\vspace{1.0cm} \LARGE \bf MERS structure

### 2017-03-06 mers-structure.tex
Add text to results and discussion.

\subsection*{Cross species transmission of MERS-CoV}
It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
We confirm this by identifying the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}).
The human side of MERS outbreaks, on the other hand, are caused by repetitive zoonotic transmissions from camels, which we estimated by using an explicit model of population structure.
We infer that across 216 MERS-CoV sequences collected from humans, we detect a median of 67 individual camel to human cross-species transmissions (95\% highest posterior density: 60--72), with little support for back-transmission scenarios (median 3, 95\% HPDs: 0--13).
There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
We also note a general predisposition for large MERS-CoV clusters to have crossed the species barrier into humans early in each year.

@@

\subsection*{Estimated epidemiological parameters}
Removing the confounding effects of shared ancestry for MERS-CoV sequences has allowed us to treat every viral introduction into humans as an independent replicate of the transmission process in humans.

@@

\subsection*{}

>>>>

\subsection*{Recombination occurs in the reservoir}
Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
Since detectable and problematic recombination in RNA viruses requires co-infection with distinct viral genotypes humans were ruled out as a suitable recombination environment \textit{a priori}.
Here we confirm this assumption by showing that evolution of MERS-CoV within human clusters is clonal.
This is not surprising, since zoonotic transmission of MERS-CoV into humans is relatively rare, of limited diversity at the point of introduction and viral diversity in humans is ultimately transient.
All of these factors act to make co-infection of humans with MERS-CoV exceedingly rare.
It is worth highlighting that MERS-CoV almost certainly recombines during human infection, but the limited diversity of virus populations in individual humans renders it invisible and of little consequence.

Although we do not detect recombination in human clusters, we see evidence for introduction of recombinant viruses into humans which evolve clonally after the transfer.
This is not unexpected either - there are strong indications that the Korean MERS outbreak was caused by a recombinant virus crossing the species barrier.
The frequency of MERS-CoV recombination in the reservoir implies that co-infection of camels is quite common, and by extension that the virus is likely to be very prevalent in this host.

\subsection*{Effects of recombination are minimal}
Most phylogenetic approaches model sequence evolution as a strictly clonal process and when recombination is present traditional phylogenetic approaches can fail to recover correct topologies, branch lengths or demographic history.
All coronavirus genera are known to be recombinogenic, and MERS-CoV is no exception, but whereas recombination is obligate for some eukaryotes, detectable RNA virus recombination can only occur during co-infection with two or more distinct viral genotypes.
Previous studies have inferred prevalent recombination in MERS-CoV and found direct evidence for it.
When not directly looking for recombination, however, most studies carried out their objectives without regard for potential effects of recombination.
Our independent analyses of the alignment split into two fragments reveals that whilst some effects of recombination are certainly detectable, they do not seem to be systematically biasing phylogenetic inference.
%So whereas we find some evidence of recombinant viruses crossing the species barrier into humans

\subsection*{Numerous cross-species transmissions of MERS-CoV}
It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
We confirm this by identifying the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}).
The human side of MERS outbreaks, on the other hand, are caused by repetitive zoonotic transmissions from camels, which we estimated by using an explicit model of population structure.
We infer that across 216 MERS-CoV sequences collected from humans, we detect a median of 67 individual camel to human cross-species transmissions (95\% highest posterior density: 60--72), with little support for back-transmission scenarios (median 3, 95\% HPDs: 0--13).
There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
We also note a general predisposition for large MERS-CoV clusters to have crossed the species barrier into humans early in each year.

@@

\subsection*{MERS epidemiology}
Although sequences in an outbreak necessarily represent a subset of all the cases, their clustering and resulting cluster sizes can be highly informative of underlying epidemiological processes that are otherwise inaccesible or unfeasible to study.
In our case we have used sequence data to disentangle the evolution of MERS-CoV into two discrete host compartments.
By removing the confounding effects of shared ancestry for MERS-CoV sequences we can treat every viral introduction into humans as an independent replicate of the transmission process in humans.

@@

\subsection*{MERS-CoV in camels}
Our observation of recombinant MERS coronaviruses being transmitted to humans is not cause for concern by itself, however, the mere presence of detectable recombination implies that co-infection of camels with MERS-CoV is common.
By extension, frequent co-infection implies a high prevalence of the virus in camels, which is the main unsettling conclusion.
There is ample evidence for this from seroepidemiology, which has shown that XX\% of camels are infected by age XX.
Our results show that zoonotic transmission of MERS-CoV into humans is the main drivers for the continuation of the outbreak in the Arabian Peninsula. 
Prevalence of MERS-CoV in camels could easily be translated into force of infection and given the role of zoonotic transmission as the main driver for the outbreak only reiterates the importance of controlling the MERS coronavirus in the reservoir if the human side of the outbreak is to be brought under control.
